Linearity and tropical mathematics in the Lambda-calculus

### 2023-07-24 CSL24/section4.tex
Commit giornaliero

A \emph{tropical polynomial} is a piece-wise linear function $f:\Lawv\to \Lawv$ of the form $f(x)=\min_{0\leq j\leq n}\{i_{j}x+c_{i_{j}}\}$, where $i_{j}\in\N$ and $c_{i_{j}}\in\Lawv$.
For example, the polynomials $\varphi_{n}(x)=\min_{0\leq j\leq n}\{jx+2^{-j}\}$
are illustrated in Fig.~\ref{fig:plot1} for $0\leq n \leq 4$.
A \emph{tropical root} of a tropical polynomials $\varphi$ is a point $x\in \Lawv$ where $\varphi$ is not differentiable. In other words, the roots of $\varphi$ are the points where the minimum defining $\varphi$ is attained at least twice (i.e.~where the slope of $\varphi$ changes).
For instance, the tropical roots of $\varphi_{n+1}$ are of the form $2^{-(i+1)}$, for $0\leq i \leq n$.
Tropical roots yield the usual factorization property of roots: if $x_{0}$ is a root of $f$, this factorizes as
$f(x)=\min\{x,x_{0}\}+ g(x)$. Yet, unlike in standard algebra, tropical roots can be computed in linear time \cite{Noferini2015}.
%Tropical polynomials and their roots are the main object of study of tropical geometry.
%
%A
% simple calculation shows that this condition corresponds, in the tropical setting, to the usual notion of root.

@@

@@

\begin{corollary}
Let $\model A$ be a finite set.
\begin{enumerate}
\item $\model{\Gamma \vdash_{\BSTLC} M:A}^!:\Lawv^{\model\Gamma} \to \Lawv^{\model A}$ is a \emph{Lipschitz} map.
\item $\model{\Gamma \vdash_{\STLC} M:A}^!:\Lawv^{\model\Gamma} \to \Lawv^{\model A}$ is a \emph{locally} Lipschitz map.
\item $\Te{M}$ decomposes $\model{\Gamma \vdash_{\STLC} M:A}^!$ as an $\inf_{t\in\Te{M}}\model{\Gamma\vdash_{\STDLC} t:A}^!$ of \emph{tropical polynomials}, thus (as $\model A$ is finite), \emph{Lipschitz} functions.
\end{enumerate}
\end{corollary} 
\begin{proof}
1). As we observed, the interpretation $\model A$ of a bounded type is a finite set, hence $\model M$ must be a finite matrix  and  its corresponding linear map $\Lawv^{\model A} \to \Lawv^{\model B}$ is thus a tropical polynomial, so we apply Corollary~\ref{prop:polylip}.
2). It follows directly from Theorem~\ref{thmTLSlocLip}.
3). It follows from the easily checked fact that, for $(f_n)_{n\in\N}\subseteq\Lawv^{!X\times Y}$, we have $\left(\inf_{n\in\N} f_n\right)^!:\Lawv^X\to\Lawv^Y$, with $\left(\inf_{n\in\N} f_n\right)^!=\inf_{n\in\N} f_n^!$.
\end{proof}

Remark that the restriction to the ground type $*$ in (2) is without loss of generality, thanks Currying.

%Moreover, from Proposition \ref{prop:boundedlip}, we can also deduce the desired Lipschitz decomposition of the Taylor expansion (using Corollary~\ref{cor:T(M)=M}):

%\begin{corollary}\label{thm:M=infLip}
% If $\Gamma \vdash_{\STLC}M:A$, then $\model M^!: \model\Gamma \to \model A$ is an $\inf$ of \emph{Lipschitz} functions (via its Taylor expansion \eqref{eq:taylorcat}).
%\end{corollary}

%A consequence of this result is that the distance between two differential programs can always be approximated via bounded programs:
%\begin{corollary}
%For all $\lambda$-terms $M,N: A\to B$ of $\STDLC$, for all $\epsilon >0$, there exists $K\in \BB N$ such that 
%$|\|\model{M}-\model{N}\|_{\infty}- |\| \model{
%\end{corollary}

>>>>

A \emph{tropical polynomial} is a piece-wise linear function $f:\Lawv\to \Lawv$ of the form $f(x)=\min_{0\leq j\leq n}\{i_{j}x+c_{i_{j}}\}$, where $i_{j}\in\N$ and $c_{i_{j}}\in\Lawv$.
For example, the polynomials $\varphi_{n}(x)=\min_{0\leq j\leq n}\{jx+2^{-j}\}$
are illustrated in Fig.~\ref{fig:plot1} for $0\leq n \leq 4$.

@@

A \emph{tropical root} of a tropical polynomials $\varphi$ is a point $x\in \Lawv$ where $\varphi$ is not differentiable. In other words, the roots of $\varphi$ are the points where the minimum defining $\varphi$ is attained at least twice (i.e.~where the slope of $\varphi$ changes).
For instance, the tropical roots of $\varphi_{n+1}$ are of the form $2^{-(i+1)}$, for $0\leq i \leq n$.
Tropical roots yield the usual factorization property of roots: if $x_{0}$ is a root of $f$, this factorizes as
$f(x)=\min\{x,x_{0}\}+ g(x)$. Yet, unlike in standard algebra, tropical roots can be computed in linear time \cite{Noferini2015}.
%Tropical polynomials and their roots are the main object of study of tropical geometry.
%
%A
% simple calculation shows that this condition corresponds, in the tropical setting, to the usual notion of root.

@@

\begin{corollary}
Let $\model A$ be a finite set.
\begin{enumerate}
\item $\model{\Gamma \vdash_{\BSTLC} M:A}^!:\Lawv^{\model\Gamma} \to \Lawv^{\model A}$ is a \emph{tropical polynomial}, thus (as $\model A$ is finite), a \emph{Lipschitz} function.
\item $\model{\Gamma \vdash_{\STLC} M:A}^!:\Lawv^{\model\Gamma} \to \Lawv^{\model A}$ is a \emph{locally} Lipschitz map.
\item $\Te{M}$ decomposes $\model{\Gamma \vdash_{\STLC} M:A}^!$ as an $\inf_{t\in\Te{M}}\model{\Gamma\vdash_{\STDLC} t:A}^!$ of \emph{tropical polynomials}, thus (as $\model A$ is finite), \emph{Lipschitz} functions.
\end{enumerate}
\end{corollary} 
\begin{proof}
1). As we observed, the interpretation $\model A$ of a bounded type is a finite set, hence $\model M$ must be a finite matrix  and  its corresponding linear map $\Lawv^{\model A} \to \Lawv^{\model B}$ is thus a tropical polynomial, so we apply Corollary~\ref{prop:polylip}.
2). It follows directly from Theorem~\ref{thmTLSlocLip}.
3). It follows from the easily checked fact that, for $(f_n)_{n\in\N}\subseteq\Lawv^{!X\times Y}$, we have $\left(\inf_{n\in\N} f_n\right)^!:\Lawv^X\to\Lawv^Y$, with $\left(\inf_{n\in\N} f_n\right)^!=\inf_{n\in\N} f_n^!$.
\end{proof}

Remark that the restriction $\model A$ finite is without loss of generality, since by Currying all programs can be seen having type $*$, which is natural to interpret as a singleton.

A consequence of (3) is that the distance between two programs can always be approximated via Lipschitz tropical polynomial approximants of the initial two programs.
\begin{corollary}
 Let $\Gamma \vdash_{\STLC} M:A$ and $\Delta\vdash_{\STLC} N:B$.
 For all $\epsilon>0$, $x\in\Lawv^{\model \Gamma}$, $b\in\model A$, there exist $t\in\Te{M}$, $u\in\Te{N}$ s.t.\ $\big| \model{\Gamma \vdash_{\STLC} M:A}^!(x)_b - \model{\Delta \vdash_{\STLC} N:B}^!(x)_b \big| \leq 2\epsilon + \big| \model{\Gamma \vdash_{\STDLC} t:A}^!(x)_b - \model{\Delta \vdash_{\STDLC} u:B}^!(x)_b \big|$.
\end{corollary}

\begin{remark}
 Let $X$, $Y$ be sets and let $\langle \_,\_\rangle:X\times Y \to \mathbb{R}$.
 For $f:X\to \mathbb R$, define $f^*:Y\to \mathbb R$ by $f^*(y):= \sup_{x\in X}\{\langle x,y\rangle - f(x)\}$.
 Then for $X=!A$, $Y=\Lawv^A$, where $A$ is a set, and $\langle \mu, y \rangle:= \mu y$, we have that $f^!=(-f)^*$ for all $f\in\Lawv^{!A}$.
 This is precisely the same formal construction yielding the well-known convex conjugate $f^*$ of a function $f$, by taking $X$ any vector space, $Y$ its dual space, and $\langle \_,\_\rangle$ the application bilinear form (acting as the scalar product on coordinates).
 This construction is in turn a generalisation of the Legendre transformation.
 Despite the formal constructions being the same, we ignore for the moment if these could be connected to the study of high-order programs in our setting.
\end{remark}

### 2023-07-24 CSL24/section5.tex
Committ
Per ora è ancora molto copia ed incolla da LICS. Fai tutte le modifiche che vuoi, se vuoi. Io domani non posso lavorare, quindi prendi il lock quando vuoi. Da domenica, dimmelo così ci organizziamo. Notte !
Commit giornaliero

In this section we illustrate a few directions in which the tropical semantics just introduced could be used to analyze quantitative properties of higher-order programs.

Since algebraic and geometric properties in tropical mathematics are usually more tractable from a computational point of view, in several well-known applications (e.g.~for optimization problems related to machine learning \cite{Pachter2004, Zhang2018, Maragos2021}) one starts from a given model, typically expressed by some polynomial function $f$, and studies 
what properties of the model can be deduced from the \emph{tropicalization} of $f$, noted $\trop f$, i.e.~the transformation of $f$ into a tropical polynomial.

%
%
%several well-known applications of tropical mathematics is to study how much can be deduced of some function starting from the properties of its tropicalization.
%In Section \ref{section5} we will follow a similar direction, investigating what quantitative properties of a higher-order programs are revealed by the study of its tropical interpretation.
%

Here we follow a similar pattern: we consider a program $M$, which can be expressed in the form of a polynomial or a power series $f$, and we 
investigate what quantitative properties of $M$ can be deduced from the properties of $\trop f$, that will indeed coincide with the interpretation of $M$ in $\LREL_{!}$.

@@

\subsection{Convergence log-probabilities}

Let us consider a probabilistic extension of $\STLC$, call it $\STLC_\oplus$:
we add a ground type $\bool$, terms $\true,\false$ of type $\bool$, terms of shape $M\oplus_p N$ and $pM$, for $p\in[0,1]$, typed via the rules:
\[
 \dfrac{\Gamma\vdash M:A \qquad \Gamma\vdash N:A}{\Gamma\vdash M\oplus_p N:A} 
 \qquad\qquad
 \dfrac{\Gamma\vdash M:A}{\Gamma\vdash pM:A}
\]
We add reduction rules:
\[
 M\oplus_p N \to pM \quad \textit{ and } \quad M\oplus_p N \to (1-p)N
\]
so that $M\oplus_p N$ plays the role of a probabilistic coin toss of bias $p$.
We use this calculus as a toy example for our purposes, and it can be seen as a fragment of the PCF considered in \cite{Manzo2013}.

Consider now the following term $M$ of type $\bool$:
\[
 (\true \oplus_p\false)\oplus_p((\true\oplus_p\false)\oplus_p(\false\oplus_p\true))
\]
Let us give addresses $\omega\in\set{00,01,100,101,110,111}$ to the occurrences of $\true,\false$ in $M$, by following the tree structure of $M$, reading $0$ as ``left'' and $1$ as ``right''.
The same addresses also represent all the different possible reduction paths from $M$ to a normal form.
For instance, $100$ represents the reduction which keeps the right part of the outermost $\oplus_p$ and erases the left part, then continues by choosing the left part twice, reaching at the end the occurrence $\true_{100}$ in $M$, i.e.\ the second occurrence of $\true$ in $M$ starting from the left.
Calling $q:=1-p$, there are the following six normal terms reachable from $M$:
$P_{00}(p,q)\true$, 
$P_{100}(p,q)\true$, 
$P_{111}(p,q)\true$, 
$P_{01}(p,q)\false$, 
$P_{110}(p,q)\false$,
$P_{101}(p,q)\false$,
where the $P$'s are the following monomials in $p,q$:
$P_{00}(p,q):=p^2$,
$P_{100}(p,q):=qp^2$,
$P_{111}(p,q):=q^3$,
$P_{01}(p,q):=pq$,
$P_{110}(p,q)=P_{101}(p,q):=q^2p$.
They correspond to the respective reduction path from $M$ to the normal term of the same address.
$P_{\omega}$ is then the probability (as a function of $p,q$) of obtaining the respective occurrence $\true_{\omega}$ or $\false_\omega$ after all the coin tossings in $M$ are performed.
Thinking of $p,q$ as parameters, $P_{\omega}$ can thus be read as the \emph{likelihood function} of the event $\omega$.
The polynomial $Q_{1}(p,q):=P_{00}(p,q)+P_{100}(p,q)+P_{111}(p,q)$ gives instead the whole probability of obtaining $\true$ after all the tossings, and $Q_{0}(p,q):=P_{01}(p,q)+P_{110}(p,q)+P_{101}(p,q)$ the one for $\false$.

@@

%These are questions much in the style of those asked in {\color{red}Ref?} in the framework of {\color{red}??}.
Answering 1) and 2) amounts at solving a maximization problem related to $P_{\omega}, Q_\omega$, which is more easily solved by 
passing to the tropical monomials/polynomial $\trop P_{\omega},\trop Q_{\omega}$. 
For 1), we are looking for $p,q\in[0,1]$ s.t.\ $q=1-p$ and $(p,q)$ belong to:
\begin{equation}
  \begin{array}{r@{}c@{}l@{}}
   \underset{(x,y)}{\RM{arg max}} P_\omega (x,y)
   & = &
   \underset{(x,y)}{\RM{arg min}}\set{-\log_c P_\omega (x,y)}
   \\
   & = &
   \underset{(x,y)}{\RM{arg max}}\set{(\trop P_\omega) (-\log_c x,-\log_c y)} \label{eq:argmax}
  \end{array}
\end{equation}
where the equalities hold for all $0<c\neq1$.

For 2), %the $\omega_0$ is s.t.\ $\max\limits_{\omega\in\set{00,100,111}} \, P_\omega(x,y) = P_{\omega_0}(x,y)$. So
we are looking for $p,q\in[0,1]$ s.t.\ $q=1-p$ and
$\max\limits_{\omega\in\set{00,100,111}} \, P_\omega(x,y) = P_{\omega_0}(x,y)$.
This last condition is equivalent to ask that:
\begin{equation}
  \begin{array}{l@{}l@{}}
    \min\limits_{\omega\in\set{00,100,111}} \, -\log_c P_\omega(x,y) = -\log_c P_{\omega_0}(x,y)
   & \textit{ i.e.: }
   \\
   (\trop Q_1)( -\log_c x, -\log_c y) = (\trop P_{\omega_0})( -\log_c x, -\log_c y). &  \label{eq:max}
  \end{array}
\end{equation}
In both cases, passing through $\trop $ %P_\omega, \trop Q_\omega$ 
makes the problem easier, as this amounts to study tropical polynomials (for instance computing tropical roots can be done in linear time \cite{Noferini2015}). %the tropicalisation operator $\trop{}$ as well as the \emph{negative $\log$-probabilities} appear.

\begin{remark}
 By adapting [Section IV, \cite{Manzo2013}], it can be seen that $\LREL_!$ is a model of $\STLC_\oplus$.
 In particular, if we set $\model \bool:=\set{0,1}$, our running example $M$ is interpreted as $\model M\in\Lawv^{\set{0,1}}$ giving the tropicalised probabilities: $\model M_0=(\trop Q_0)(p,1-p)$, $\model M_1=(\trop Q_1)(p,1-p)$.
\end{remark}

%Now, in $\LREL$, seen as a model of such probabilistic $\lam$-calculus, the interpretation of a term already computes the tropicalisation of the polynomials expressing the probabilities, because the underlying semiring of the model is tropical.
%For instance, for our running example $M$:
%\[\model M = \min\left\{\trop{Q_1}(p,1-p) \cdot \model I, \trop{Q_0}(p,1-p) \cdot \model \Omega\right\}.\]
Therefore, assuming the interpretation of a term is known, 
so that one can extract the associated tropical polynomials from it,
% (in $p$ and $q$, not after the substitution $q:=1-p$) 
% are extracted from it, 
one could study the optimisation problems \ref{eq:argmax}, \ref{eq:max} \emph{without} having to actually reduce the term.

%\begin{example}\label{ex:study}
For our running example $M$, we have $\trop Q_1(x,y)=\min\set{2x,y+2x,3y}$ and $\trop Q_0(x,y)=\min\set{x+y,2y+x}$.
Studying $\trop Q_1$, whose plot is in Fig.~\ref{fig:plot2}, we see that $\trop Q_1(x,y)=3y$ iff $y\leq \frac{2}{3}x$, and it coincides with $2x$ otherwise.
Remembering that $3y=P_{111}(x,y)$, we can now solve the optimisation problem~\ref{eq:max} for $\omega_0=111$:
via the substitution $x:=-\log_c p$, $y:=-\log_c (1-p)$, Equation~\ref{eq:max} is equivalent to $-\log_c (1-p)\leq -\frac{2}{3}\log_c p$, i.e.\ $1-p\geq p^{\frac{2}{3}}$.
This means that, for $p\in[0,1]$ s.t.\ $1-p\geq p^{\frac{2}{3}}$ (for example, $p=\frac{1}{4}$), the most likely occurrence of $\true$ to obtain, knowing that $M$ sampled $\true$ in its normal form, is $\true_{111}$.
Remembering that $2x=P_{00}(x,y)$, for the other values of $p$ (for example, $p=\frac{1}{2}$), the most likely $\true$ to be sampled is the occurrence $\true_{00}$.
We have thus answered question 2) above.
%\end{example}

>>>>

%In this section we illustrate a few directions in which the tropical semantics just introduced could be used to analyze quantitative properties of higher-order programs.

Since algebraic and geometric properties in tropical mathematics are usually more tractable from a computational point of view, in several well-known applications (e.g.~for optimization problems related to machine learning \cite{Pachter2004, Zhang2018, Maragos2021}) one starts from a given model, typically expressed by some polynomial function $f$, and studies 
what properties of the model can be deduced from the \emph{tropicalization} of $f$, noted $\trop f$, i.e.~the transformation of $f$ into a tropical polynomial.
%
%
%several well-known applications of tropical mathematics is to study how much can be deduced of some function starting from the properties of its tropicalization.
%In Section \ref{section5} we will follow a similar direction, investigating what quantitative properties of a higher-order programs are revealed by the study of its tropical interpretation.
%
Here we follow a similar pattern: we consider a program $M$, which can be expressed in the form of a polynomial or a power series $f$, and we 
investigate what quantitative properties of $M$ can be deduced from the properties of $\trop f$, that will indeed coincide with the interpretation of $M$ in $\LREL_{!}$.

@@

\end{comment}

\subsection{Log-probabilities with a worked out example}

Let us consider a probabilistic extension of $\STLC$, call it $\STLC_\oplus$:
we add a ground type $\bool$, terms $\true,\false$ of type $\bool$, terms of shape $M\oplus_p N$ and $pM$, for $p\in[0,1]$, typed via the usual rules and whose operational semantics adds the usual rules:
{\small{\[\begin{array}{cccc}
           \dfrac{\Gamma\vdash M:A \qquad \Gamma\vdash N:A}{\Gamma\vdash M\oplus_p N:A}
           &
           \dfrac{\Gamma\vdash M:A}{\Gamma\vdash pM:A}
           &
           M\oplus_p N \to pM
           &
           M\oplus_p N \to (1-p)N
          \end{array}\]}}
%We add reduction rules:$ M\oplus_p N \to pM \quad$ and $\quad M\oplus_p N \to (1-p)N,$
so that $M\oplus_p N$ plays the role of a probabilistic coin toss of bias $p$.
We use this calculus as a toy example for our purposes.
It can be compiled into $\mathrm{PCF}^{\Lawv}$, an instance of the $\mathrm{PCF}^{\mathcal R}$-languages studied in \cite{Manzo2013}, via the translation $(\_)^\circ$ of terms generated by $(M\oplus_p N)^\circ:=pM^\circ\prog{ or } (1-p)N^\circ$.
It is easy to see that if $\Gamma \vdash_{\STLC_\oplus} M:A$, then $\Gamma \vdash_{\mathrm{PCF}^{\Lawv}} M^\circ:A$.
This is slightly imprecise because in \cite{Manzo2013}, their $\mathrm{PCF}^{\mathcal R}$-languages only have one ground type $\mathrm{int}$, used for modeling $\N$, and has no term $\prog{True}$, $\prog{False}$.
So we are actually compiling $\STLC_\oplus$ into the immediate extension of $\mathrm{PCF}^{\mathcal R}$ which adds two new ground types, $*$ and $\bool$ (and, since we will not consider the integers, we can also drop $\mathrm{int}$), plus the two boolean terms.
This does not invalidate the results (not concerning $\mathrm{int}$) of \cite{Manzo2013}.
Remark also that the operational semantics of $\mathrm{PCF}^{\Lawv}$ \cite[Fig.\ 1]{Manzo2013} simulates the one of $\STLC_\oplus$, potentially using multiple steps, e.g.\ $(M\oplus_p N)^\circ \overset{0}{\rightarrow} pM^\circ \overset{p}{\rightarrow} M^\circ$.
Since $\LREL_!$ is an instance of models for $\mathrm{PCF}^{\mathcal{R}}$, similarly to \cite[Section VI]{Manzo2013}, we use it in order to get a model of $\STLC_\oplus$, by setting $\model{\Gamma \vdash_{\STLC_\oplus} M:A}:=\model{\Gamma \vdash_{\mathrm{PCF}^{\Lawv}} M^\circ:A}\in\HOM{\LREL_!}{\model{\Gamma}}{\model{A}}$.
Here we set $\model{\bool}:=\{0,1\}$ and $\model{\vdash_{\mathrm{PCF}^{\Lawv}} \true:\bool}\in\Lawv^{\{0,1\}}$ gives value $0$ on $1$ and value $\infty$ on $0$, similarly for $\false$.

\begin{remark}
 Now the matrix interpreting a term need no longer be boolean.
 For instance, $\model{\vdash_{\STLC_\oplus} \true:\bool}_i$ is $p$ if $i=1$ and it is $1-p$ if $i=0$.
 In general, one can show that $\model{\Gamma \vdash_{\STLC_\oplus} M:A}\in[0,1]^{!\model \Gamma \times \model A}$.
 Coefficients $> 1$ may be considered when dealing with other kinds of effects.
\end{remark}

Consider %the following closed term $M$ of type $\bool$:
$
 \vdash_{\STLC_\oplus} M:=(\true \oplus_p\false)\oplus_p((\true\oplus_p\false)\oplus_p(\false\oplus_p\true)):\bool.
 $
Let us give addresses $\omega\in\set{00,01,100,101,110,111}$ to the occurrences of $\true,\false$ in $M$, by following the tree structure of $M$, reading $0$ as ``left'' and $1$ as ``right''.
The same addresses also represent all the different possible reduction paths from $M$ to a normal form.
%For instance, $100$ represents the reduction which keeps the right part of the outermost $\oplus_p$ and erases the left part, then continues by choosing the left part twice, reaching at the end the occurrence $\true_{100}$ in $M$, i.e.\ the second occurrence of $\true$ in $M$ starting from the left.
Calling $q:=1-p$, there are the following six normal terms reachable from $M$:
$P_{00}(p,q)\true$, 
$P_{100}(p,q)\true$, 
$P_{111}(p,q)\true$, 
$P_{01}(p,q)\false$, 
$P_{110}(p,q)\false$,
$P_{101}(p,q)\false$,
where the $P$'s are the following monomials in $p,q$:
$P_{00}(p,q):=p^2$,
$P_{100}(p,q):=qp^2$,
$P_{111}(p,q):=q^3$,
$P_{01}(p,q):=pq$,
$P_{110}(p,q)=P_{101}(p,q):=q^2p$.
They correspond to the respective reduction path from $M$ to the normal term of the same address.
$P_{\omega}$ is then the probability (as a function of $p,q$) of obtaining the respective occurrence $\true_{\omega}$ or $\false_\omega$ after all the coin tossings in $M$ are performed.
Thinking of $p,q$ as parameters, $P_{\omega}$ can thus be read as the \emph{likelihood function} of the event $\omega$.
The polynomial $Q_{1}(p,q):=P_{00}(p,q)+P_{100}(p,q)+P_{111}(p,q)$ gives instead the whole probability of obtaining $\true$ after all the tossings, and $Q_{0}(p,q):=P_{01}(p,q)+P_{110}(p,q)+P_{101}(p,q)$ the one for $\false$.

@@

%These are questions much in the style of those asked in {\color{red}Ref?} in the framework of {\color{red}??}.
Answering 1) and 2) amounts at solving a maximization problem related to $P_{\omega}, Q_\omega$, which is more easily solved by 
passing to the tropical monomials/polynomial $\trop P_{\omega},\trop Q_{\omega}$. 
For 1), we are looking for $p,q\in[0,1]$ s.t.\ $q=1-p$ and $(p,q)\in
%\begin{equation}
%  \begin{array}{ccccc}
   \underset{(x,y)}{\RM{arg max}} P_\omega (x,y)
   %& 
   = %&
   \underset{(x,y)}{\RM{arg min}}\set{-\log_c P_\omega (x,y)}
   %&
   =%&
   \underset{(x,y)}{\RM{arg max}}\set{(\trop P_\omega) (-\log_c x,-\log_c y)} \label{eq:argmax}$
%  \end{array}
%\end{equation}
where the equalities hold for all $0<c\neq1$.
For 2), %the $\omega_0$ is s.t.\ $\max\limits_{\omega\in\set{00,100,111}} \, P_\omega(x,y) = P_{\omega_0}(x,y)$. So
we are looking for $p,q\in[0,1]$ s.t.\ $q=1-p$ and
$\max\limits_{\omega\in\set{00,100,111}} \, P_\omega(x,y) = P_{\omega_0}(x,y)$.
This last condition is equivalent to ask that $\min\limits_{\omega\in\set{00,100,111}} \, -\log_c P_\omega(x,y) = -\log_c P_{\omega_0}(x,y)$, i.e.\ $(\trop Q_1)( -\log_c x, -\log_c y) = (\trop P_{\omega_0})( -\log_c x, -\log_c y)\label{eq:max}$.
In both cases, passing through $\trop $ %P_\omega, \trop Q_\omega$ 
makes the problem easier, as this amounts to study tropical polynomials (for instance computing tropical roots can be done in linear time \cite{Noferini2015}). %the tropicalisation operator $\trop{}$ as well as the \emph{negative $\log$-probabilities} appear.

\begin{remark}
 The interpretation $\model{\vdash_{\STLC_\oplus} M:\bool}\in\Lawv^{\set{0,1}}$ of our running example gives the tropicalised probabilities: $\model M_0=(\trop Q_0)(p,1-p)$, $\model M_1=(\trop Q_1)(p,1-p)$.
 Therefore, the model $\LREL_!$ expresses such optimisation problems 
% (in $p$ and $q$, not after the substitution $q:=1-p$) 
% are extracted from it,
 \ref{eq:argmax}, \ref{eq:max} of $\STLC_\oplus$-programs. 
\end{remark}

%Now, in $\LREL$, seen as a model of such probabilistic $\lam$-calculus, the interpretation of a term already computes the tropicalisation of the polynomials expressing the probabilities, because the underlying semiring of the model is tropical.
%For instance, for our running example $M$:
%\[\model M = \min\left\{\trop{Q_1}(p,1-p) \cdot \model I, \trop{Q_0}(p,1-p) \cdot \model \Omega\right\}.\]

%\begin{example}\label{ex:study}
For our running example $M$, we have $\trop Q_1(x,y)=\min\set{2x,y+2x,3y}$ and $\trop Q_0(x,y)=\min\set{x+y,2y+x}$.
Studying $\trop Q_1$ %, whose plot is in Fig.~\ref{fig:plot2}, 
we see that $\trop Q_1(x,y)=3y$ iff $y\leq \frac{2}{3}x$, and it coincides with $2x$ otherwise.
Remembering that $3y=P_{111}(x,y)$, we can now solve the optimisation problem~\ref{eq:max} for $\omega_0=111$:
via the substitution $x:=-\log_c p$, $y:=-\log_c (1-p)$, Equation~\ref{eq:max} is equivalent to $-\log_c (1-p)\leq -\frac{2}{3}\log_c p$, i.e.\ $1-p\geq p^{\frac{2}{3}}$.
This means that, for $p\in[0,1]$ s.t.\ $1-p\geq p^{\frac{2}{3}}$ (for example, $p=\frac{1}{4}$), the most likely occurrence of $\true$ to obtain, knowing that $M$ sampled $\true$ in its normal form, is $\true_{111}$.
Remembering that $2x=P_{00}(x,y)$, for the other values of $p$ (for example, $p=\frac{1}{2}$), the most likely $\true$ to be sampled is the occurrence $\true_{00}$.
We have thus answered question 2) above.
%\end{example}
